---
title: "Forest_Fire_Initial"
author: "Debalina Maiti"
date: "5/24/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis is motivated by the following research question:

\begin{center}
What factors lead to particularly damaging forest fires?
\end{center}

## Setup

First, we load the car library, which gives us a convenient scatterplotMatrix function.

```{r car}
library(car)
```

```{r loading}
# Load the data 
setwd("/Users/debalinamaiti/Documents/GraduateStudy/MIDS/W203/Unit4/HomeWork/Forest_Fire_EDA")

read.csv("forestfires.csv")
forest_fire = read.table("forestfires.csv", header=TRUE, sep=",", na.string = "na")
```

## Data Overview


```{r shape}
nrow(forest_fire)
str(forest_fire)
forest_fire
```

# Data Arrangement
===================
We will try to make different type of data transformations for our analysis

#ff_level
We will add a new column called 'level', to the data frame. And will name this new data frame as ff_level
This level column will contain the value: high, medium and low
For the rows with area = 0, we will populate the level value as 'low'
For the rows with 0 < area < 12.85, we will populate the level value as 'medium'
For the rows with area >= 12.85, we will populate the level value as 'high'
#--

```{r}

## Case High

case_high = subset(forest_fire, 12.85<=forest_fire$area &!is.na(forest_fire$area))
case_high["level"]<- "high"

nrow(case_high)##high count = 76

## Case Medium

case_medium = subset(forest_fire, 0<forest_fire$area & forest_fire$area<12.85 &!is.na(forest_fire$area))
case_medium["level"]<- "medium"

nrow(case_medium)##medium count = 194

## Case Low

case_low = subset(forest_fire, forest_fire$area==0 &!is.na(forest_fire$area))
case_low["level"]<- "low"

nrow(case_low)##low count = 247

###Merge all three to ff_level

ff_level = merge(case_high, case_medium, all = TRUE)
ff_level = merge(case_low, ff_level, all = TRUE)

####Logerithm - looking at the data against area we see that there are huge number of data with value zero. So will replace area value with log(area) value
#
ff_level$area = log(ff_level$area)
str(ff_level)
```

#Plot X, Y 
```{r}
plot(jitter(ff_level$X, factor=2),jitter(ff_level$Y, factor=2 ), col = factor(ff_level$level))
legend(1, legend=levels(factor(ff_level$level)), col= factor(ff_level$level))
legend=levels(ff_level$level)
```
