---
title: "Forest_Fire_Initial"
author: "Debalina Maiti, Mark Paluta, Tina Agarwal, Vivek Agarwal"
date: "5/24/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis is motivated by the following research question:

\begin{center}
What factors lead to particularly damaging forest fires?
\end{center}

## Setup

First, we load the car library, which gives us a convenient scatterplotMatrix function.

```{r car}
library(car)
```

```{r, results='hide'}
# Load the data 
read.csv("forestfires.csv")
forest_fire = read.table("forestfires.csv", header=TRUE, sep=",", na.string = "na")
```

## Data Overview


```{r shape}
nrow(forest_fire)
str(forest_fire)
# forest_fire
```

# Data Arrangement
===================
We will try to make different type of data transformations for our analysis

#ff_level
We will add a new column called 'level', to the data frame. And will name this new data frame as ff_level
This level column will contain the value: high, medium and low
For the rows with area = 0, we will populate the level value as 'low'
For the rows with 0 < area < 12.85, we will populate the level value as 'medium'
For the rows with area >= 12.85, we will populate the level value as 'high'
#--

```{r}

## Case High

case_high = subset(forest_fire, 12.85<=forest_fire$area &!is.na(forest_fire$area))
case_high["level"]<- "high"

nrow(case_high)##high count = 76

## Case Medium

case_medium = subset(forest_fire, 0<forest_fire$area & forest_fire$area<12.85 &!is.na(forest_fire$area))
case_medium["level"]<- "medium"

nrow(case_medium)##medium count = 194

## Case Low

case_low = subset(forest_fire, forest_fire$area==0 &!is.na(forest_fire$area))
case_low["level"]<- "low"

nrow(case_low)##low count = 247

###Merge all three to ff_level

ff_level = merge(case_high, case_medium, all = TRUE)
ff_level = merge(case_low, ff_level, all = TRUE)

####Logerithm - looking at the data against area we see that there are huge number of data with value zero. So will replace area value with log(area) value
#
ff_level$area = log(ff_level$area)
str(ff_level)
```

<!-- #Plot X, Y  -->
<!-- ```{r} -->
<!-- plot(jitter(ff_level$X, factor=2),jitter(ff_level$Y, factor=2 ), col = factor(ff_level$level)) -->
<!-- legend(1, legend=levels(factor(ff_level$level)), col= factor(ff_level$level)) -->
<!-- legend=levels(ff_level$level) -->
<!-- ``` -->

# Area

Area appears to be right-skewed. A log transformation helps normalize the data.

```{r}
forest_fire$logArea = log(forest_fire$area)
hist(forest_fire$area)
hist(forest_fire$logArea)
```

# Spatial coordinates

The X and Y coordinates have no missing values.

```{r}
sum(is.na(forest_fire$X))
sum(is.na(forest_fire$Y))
```

# test

Plotting the coordinates on a grid, we can see that some areas of the grid are densely populated with data and other areas are sparse. The colors represent size of the fire.

```{r}
# filler - define a factor for category of fire
forest_fire$fire_size <- cut(forest_fire$area, breaks=3, labels=c('Low','Medium','High'))
plot(jitter(forest_fire$X), jitter(forest_fire$Y), col=forest_fire$fire_size)
legend(1,9,unique(forest_fire$fire_size),col=1:length(forest_fire$fire_size),pch=1)
```

# Month

A histogram shows that most of the data comes from spring and summer months. We can use box plots of month and log of area to see the distributions of burned area by month.

```{r}
forest_fire$month = factor(forest_fire$month,levels(forest_fire$month)[c(5,4,8,1,9,7,6,2,12,11,10,3)])
barplot(table(forest_fire$month))
library(ggplot2)
ggplot(data = forest_fire, aes(x = month, y = logArea)) + geom_boxplot(aes(fill = logArea), width = 0.8) + theme_bw()
```


## Introduction

In some areas, forest fires are a major environmental concern, endangering human lives and causing substantial
economic damage. This analysis is motivated by the following research question:What factors lead to particularly damaging forest fires? 

## Setup
First we load the car library which gives us a convenient scatterplotMatrix function.
```{r car}
library(car)
```

#Data Selection

```{r}
#We note that we have 517 observations and 13 variables
nrow(forestfires)

```
```{r}
str(forestfires)
```

Looking at the rain and area variable, we notice that there are mostly 0's.

# Look at the summary of statistics for the forestfires
```{r}
summary(forestfires)
```
#Lets look at the summary of area 
```{r}
summary(forestfires$area)
```

Area seems to be right skewed. So to normalize the data, we are taking log of area

```{r}
forestfires$logarea = log(forestfires$area)
hist(forestfires$area)
hist(forestfires$logarea)
```

#Exploratory Analaysis

We have a large number of variables, we begin with a scatterplot matrix. This is helpful for getting
a high-level overview of the relationships between our variables and can draw our attention to important
features we want to investigate further.

```{r}
#scatterplotMatrix(~ RH + wind + rain + logarea, data=forestfires, smooth = TRUE, main = "Scatterplot Matrix for Relative Humidity, Wind and Rain")
```


#create histogram of wind
```{r}
summary(forestfires$wind)

hist(forestfires$wind,breaks=seq(0.4,9.4,0.5),main = "Wind Speed",xlab= NULL)
```
Visually, the histogram seems to have a positive skew (right skew). This means that there are
observations stretching further to the right of the bulk of the data. Note from summary that the mean is greater
than the median, which is typically what we see for positively skewed variables.

#create histogram of Relative hUmidity
```{r}
summary(forestfires$RH)
hist(forestfires$RH,breaks=seq(15,100,5),main = "Percentage of Relative Humidity",xlab= "RH in %")
```
Visually, the histogram seems to have a positive skew (right skew) again. This means that there are
observations stretching further to the right of the bulk of the data. Note from summary that the mean is greater
than the median, which is typically what we see for positively skewed variables.

We next examine our Rain variable

#create histogram of Rain 
```{r}
summary(forestfires$rain)
hist(forestfires$rain, breaks=seq(0.0,6.4,0.2), main = "Outside rain in mm/m2", xlab = "Rain in mm/m2" )
```



It almost seems like that distribution is spiked at around zero, which means that there was no rain at all.

#boxplot
```{r}
#boxplot(rain ~ RH ~ logarea, data = forestfires)
```
